Efficient pedigree recording for fast population genetics simulation

### 2018-01-15 forwards_paper.tex
minor fixen

%% Memory reduction thanks to simplification
First: how much memory do simplified tree sequences require?
Consider a simulation of a Wright--Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and $2NT$ edges.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals, back to either $T$ generations ago,
or their common ancestor, whichever comes first.
The tree sequence must store edges describing the leftmost marginal tree,
which requires at most $N-2$ edges.
Then, each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Suppose that $\mathcal{T}_x$ is the marginal tree at genomic position $x$,
and write $|\mathcal{T}_x|$ for the total length of the tree.
For the tree at nearby position $x + dx$ to be different,
there must have been a crossing-over between $x$ and $x + dx$
in one of the $|\mathcal{T}_x|$ meioses that gave birth to those individuals.
(Recall that these ``individuals'' are haploid.)
Since we measure distance along the genome
so that length is equal to the expected number of crossing-overs per generation,
the expected distance until the next crossing-over is $1/|\mathcal{T}_x|$.
If every crossing-over changed the marginal tree,
then this would imply, for consistency,
that the expected total number of times that the marginal tree changes along the genome
is equal to the total area of the tree sequence (as defined above).
Not every such crossing-over changes the marginal tree, but most do,
so the total area of the tree sequence,
multiplied by four,
gives an upper bound on the expected number of edges in the tree sequence
beyond those describing the leftmost tree.
Since we are considering a chromosome of length 1,
the expected total area is equal to the mean marginal tree length, as above.

@@

%% Progressive memory reduction as a function of T
What if $T$ is smaller:
how many of the resulting $2NT$ edges are required after simplification?
In other words, how fast does the information in the pedigree become irrelevant?
% \citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
% by retaining only 8 generations of history.
% This begs the question: how fast does the information in the pedigree become irrelevant?
Now, we need to compute the expected total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Again, coalescent theory tells us that
the expected length of time for which a coalescent tree has $k$ lineages
is $2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since the tree has $k$ branches over this period,
it is expected to contribute $2N/(k-1)$ to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2(N/n - 1)$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
% since 2N(1/n - 1/N) = 2(N/n-1) = T implies
%    N/n = (T+2)/2 and so n = 2N/(T+2)
is expected to have around $r(T)$ roots, where $r(T) \approx 2N/(T+2)$.
The total tree length over this time is
$2N \sum_{k=N}^{r(T)+1} 1/(k-1) \approx 2 N \log(N/r(T))$.
This is a crude estimate for several reasons:
first, we should not count the branch leading to the root of the tree (i.e., when $r(T)=1$),
and second, this does not account for the discrete nature of the Wright--Fisher model.
% which is approximately
% \begin{align*}
%     2N - 2 + 8 N \left\{ \log(N-1) - \log(r(T)) \}
%        = 2N - 2 + 8 N \log( (T+2)(N-1)/2N )
%        \approx 2 N (1 + 4 \log( (T+2)/2 ))
% \end{align*}
Nonetheless, this leads as above to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \min\left(N, \frac{T+2}{2}\right) \right)\right) .
\end{align}
This implies that the number of edges required to store the last $T$ generations
of history for the entire chromosome of a population of size $N$
grows as $O(N \log T)$ -- proportionally to $N$ at first but rapidly tapering off.
The bound holds up reasonably well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.

>>>>

%% Memory reduction thanks to simplification
First: how much memory do simplified tree sequences require?
Consider a simulation of a Wright--Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and $2NT$ edges.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals, back to either $T$ generations ago,
or their common ancestor, whichever comes first.
The tree sequence must store edges describing the leftmost marginal tree,
which requires at most $2N-2$ edges.
Then, each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Suppose that $\mathcal{T}_x$ is the marginal tree at genomic position $x$,
and write $|\mathcal{T}_x|$ for the total length of the tree.
For the tree at nearby position $x + dx$ to be different,
there must have been a crossing-over between $x$ and $x + dx$
in one of the $|\mathcal{T}_x|$ meioses that gave birth to those individuals.
(Recall that these ``individuals'' are haploid.)
Since we measure distance along the genome
so that length is equal to the expected number of crossing-overs per generation,
the expected distance until the next crossing-over is $1/|\mathcal{T}_x|$.
If every crossing-over changed the marginal tree,
then this would imply, for consistency,
that the expected total number of times that the marginal tree changes along the genome
is equal to the total area of the tree sequence (as defined above).
Not every such crossing-over changes the marginal tree, but most do,
so the total area of the tree sequence,
multiplied by four,
gives an upper bound on the expected number of edges in the tree sequence
beyond those describing the leftmost tree.
Since we are considering a chromosome of length 1,
the expected total area is equal to the mean marginal tree length, as above.

@@

%% Progressive memory reduction as a function of T
What if $T$ is smaller:
how many of the resulting $2NT$ edges are required after simplification?
In other words, how fast does the information in the pedigree become irrelevant?
% \citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
% by retaining only 8 generations of history.
% This begs the question: how fast does the information in the pedigree become irrelevant?
Now, we need to compute the expected total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Again, coalescent theory tells us that
the expected length of time for which a coalescent tree has $k$ lineages
is $2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since the tree has $k$ branches over this period,
it is expected to contribute $2N/(k-1)$ to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2(N/n - 1)$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
% since 2N(1/n - 1/N) = 2(N/n-1) = T implies
%    N/n = (T+2)/2 and so n = 2N/(T+2)
is expected to have around $r(T)$ roots, where $r(T) = 2N/(T+2)$.
The total tree length over this time is
$2N \sum_{k=r(T)+1}^N 1/(k-1) \approx 2 N \log(N/r(T))$.
This is a crude estimate for several reasons:
first, we should not count the branch leading to the root of the tree (i.e., when $r(T)=1$),
and second, this does not account for the discrete nature of the Wright--Fisher model.
% which is approximately
% \begin{align*}
%     2N - 2 + 8 N \left\{ \log(N-1) - \log(r(T)) \}
%        = 2N - 2 + 8 N \log( (T+2)(N-1)/2N )
%        \approx 2 N (1 + 4 \log( (T+2)/2 ))
% \end{align*}
Nonetheless, this leads as above to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \min\left(N, \frac{T+2}{2}\right) \right)\right) .
\end{align}
This implies that the number of edges required to store the last $T$ generations
of history for the entire chromosome of a population of size $N$
grows as $O(N \log T)$ -- proportionally to $N$ at first but rapidly tapering off.
The bound holds up reasonably well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
